---
title: "Project2"
author: "Emma Aldred efa364"
date: "11/22/2020"
output: 
  html_document:
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
  pdf_document:
    toc: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse);library(lmtest);library(plotROC);library(glmnet)
library(sandwich);library(readr);library(rstatix)
```

Introduction
  
  This project analyzes variables within the dataset "Guns" uploaded as a CSV from https://vincentarelbundock.github.io/Rdatasets/datasets.html. The Guns dataset has data from 1977-1999 in all 50 US states and the District of Columbia. The data collected includes "violent" crime rate, "murder" rate, "robbery" rate, and "prisoners" - incarceration rate from the previous year - all per 100,000 members of the population. It also have demographic variables: "afam" is the percent of the state population (ages 10-64) that is African American, "cauc" is the percent of the state population (ages 10-64) that is Caucasian, "male" is the percent of the state population (ages 10-29) that is male, state "population" in millions of people, mean personal "income" in US dollars, and population "density". Finally, the dataset includes a variable "law" which is a yes/no variable for whether or not the state has a shall carry law in effect for the year. Each state + DC has a value for each of the variables listed above for all the years between 1977-1999 making a total of 1,172 observations.

```{r cars}
Guns<- read.csv("Guns.csv")
head(Guns)
```

MANOVA Test
```{r pressure, echo=FALSE}
group <- Guns$law 
DVs <- Guns %>% select(violent,murder,robbery,prisoners,afam,cauc,male,population,income,density)

#Test multivariate normality for each group (null: assumption met)
sapply(split(DVs,group), mshapiro_test) #significant therefore assumption failed

#If any p<.05, stop. If not, test homogeneity of covariance matrices

Guns_man<-manova(cbind(violent,murder,robbery,prisoners,afam,cauc,male,population,income,density)~law,data=Guns) 
summary(Guns_man) #MANOVA is significant 

summary.aov(Guns_man) #Violent, murder, robbery, afam, cauc, male, population, density are signficant

Guns%>%group_by(law)%>%summarize(mean(violent),mean(murder),mean(robbery),mean(afam),mean(cauc),mean(male),mean(population),mean(density))

#1 manova, 10 anova= 11 tests
1-.95^11
#probability of at least 1 type one error (if p unadjusted) = 0.4311999
.05/11
#p=0.004545455
#all still signficant
```
  The MANOVA test was done on all numeric variables in the dataset (year is considered a categorical label). The assumptions for MANOVA were not met and, therefore, I am skeptical about the finding from this test. Regardless, the MANOVA test was conducted to determine the effect of the variable law on various crime statistics (violent, murder, robbery, prisoner) and demographics (afam, cauc, male, population, and density). 
  Significant differences were found among law for at least one of the dependent variables (Pillai trace= 0.23, psuedo F(10,1162)=35.474, p<.0001). 10 univariate ANOVAs were performed as a follow up and were significant for violent (F(1,1171)=52.364,p<.0001), murder (F(1,1171)=38.944,p<.0001), robbery (F(1,1171)=55.363,p<.0001), afam (F(1,1171)=41.009,p<.0001), cauc (F(1,1171)=55.29,p<.0001), male (F(1,1171)=97.477,p<.0001), population (F(1,1171)=18.397,p<.0001), and density (F(1,1171)=15.04,p<.0001). A post hoc analysis did not need to be perfomed since there are only two categories for the law variable (yes/no). Law - yes or no - was found to differ significantly from in terms of violent, murder, robbery, afam, cauc, male, population, and density. Since a total of 11 tests were conducted (1 manova and 10 anova), the Bonferroni method was used to control for Type 1 error rates for multiple comparisons. Without this correction, the probability of at least one 1 type error would be .43; Therefore, a p-value of 0.0045 was used, but this correction did not change the significance of any off the variables.

Randomization Test
```{R}
#Ho: mean violent crime rate is the same for states with and without shall carry laws
#Ha: mean violent crime rate is different for states with and without shall carry laws
Guns%>%group_by(law)%>%summarize(Avg=mean(violent))%>%summarize(diff(Avg))

rand_crime<-vector()
for(i in 1:5000){
new<-data.frame(vcrime=sample(Guns$violent),carry_law=Guns$law)
rand_crime[i]<- mean(new[new$carry_law=="yes",]$vcrime)-mean(new[new$carry_law=="no",]$vcrime)
}
mean(rand_crime< -161.1868|rand_crime> 161.1868) #reject Ho

{hist(rand_crime,main="Null Distribution of Random Crime Rate",ylab="Frequency",xlim = c(-162,162)); abline(v = c(-161.1868, 161.1868),col="red")}
```
  
  The randomization test was performed to determine if the following hypotheses were true: 
Ho: mean violent crime rate is the same for states with and without shall carry laws.
Ha: mean violent crime rate is different for states with and without shall carry laws.
A p value of 0.00 was computed which is less that .001 and, therefore, the null hypothesis can be rejected. There is evidence that the mean violent crime rate is different for states with and without shall carry laws

Linear Regression Model
```{R}
Guns$y<-ifelse(Guns$law=="yes",1,0)
Guns$prisoners_c<-Guns$prisoners-mean(Guns$prisoners)
Guns$violent_c<-Guns$violent-mean(Guns$violent)
fit<-lm(violent_c~y*prisoners_c, data=Guns)
summary(fit)
ggplot(Guns, aes(prisoners_c,violent_c, color = law)) + geom_smooth(method = "lm", se = F, fullrange = T)+geom_point()+ggtitle("State Violent Crime Rate by Prisoner Incarceration Rate and 'Shall Carry Law'")+xlab("Mean Incarceration Rate")+ylab("Mean Violent Crime Rate")

#Assumption Checking LINE
ggplot(Guns,aes(violent_c,prisoners_c))+geom_point() #linear relationship assumption
resids<-fit$residuals
fitted<-fit$fitted.values
shapiro.test(resids) #not normal
ggplot()+geom_histogram(aes(resids), bins=50)
bptest(fit) #variance not constant
ggplot()+geom_point(aes(fitted,resids))+geom_hline(yintercept=0, color='red')

coeftest(fit,vcov=vcovHC(fit))[,1:2]

```
  The lm model was constructed to investigate the effects of prisoners incarceration rate and the shall carry law on violent crime rates across the US States + DC. For states with average prisoner incarceration rates, the predicted violent crime rates were 182.85392 lower if the state had a 'shall carry' law. For states without the 'shall carry' law, the predicted violent crime rates is 1.36556 higher for every unit increase in prisoner incarceration rate. The slope of mean prisoner incarcerataion rate on violent crime is 0.18382 less for states without the 'shall carry law'.
  The linear fit model showed a fairly linear relationship between mean violent crime and mean prisoner incarceration rate. The Shapiro-Wilk test in which Ho = there is normally distributed residuals yeilded a p-value of < 2.2e-16 and thus the normality assumption failed. Furthermore, a Breusch-Pagan test was performeed in which Ho = there is equal variance of points/residuals along regression line yeilded a p value < 2.2e-16 and thus the homoskedasticity assumption also failed to be met. In the histogram of the residuals and in the scatterplot of the fitted values and residuals, the majority of the data is clustered together with a smaller - yet noticeable - portion of the data much further out. I predict that the smaller portion of the dataset that deviates from the majority is the likely reason the assumptions failed to be met.
  The robust standard errors of the linear model was computed and the output resulted in the same coefficients but a higher standard error (and thus a higher p value) for the intercept, prisoners_c and the interaction between y and prisoners_c and a lower standard error (and thus lower p value) for y. This model is more resistant to the lack of homoskedasticity than the original model. 55.13% of the variation in mean violent crime is being explained by the linear relationship with mean prisoner incarceration rate and the presence or lack thereof a 'shall carry' law.

Bootstrapped Standard Errors - Resampling the Residuals
```{R}
resid_crimesamp<-replicate(5000,{
new_resids<-sample(resids,replace=TRUE) 
Guns$new_violent<-fitted+new_resids 
fit<-lm(new_violent~y*prisoners_c,data=Guns)
coef(fit)
})
resid_crimesamp%>%t%>%as.data.frame%>%summarize_all(sd)
```
Bootstrapping the residuals on the same regression model resulted in all standard errors that were less (and thus p values that were less) than the original model and very similar to the robust standard error output. The bootstrapping standard erros were slighlty lower compared to the robust SE output for the intercept (7.455181), prisoners_c (0.04031502), and the interaction between prisoners_c and y (0.09646994) but higher for the dummy variable y for 'shall carry' law (15.17598).

Binary Logistic Regression Model - Prisoners and Violent Crime
```{R}
fit2<-glm(y~prisoners_c+violent_c, data=Guns,family="binomial")
summary(fit2)
Guns$prob<-predict(fit2,type="response")
Guns$pred<-ifelse(Guns$prob>.5,1,0)
table(truth=Guns$y, prediction=Guns$pred>.5)%>%addmargins

class_diag<-function(probs,truth){
  tab<-table(factor(probs>.5,levels=c("FALSE","TRUE")),truth)
  acc=sum(diag(tab))/sum(tab)
  sens=tab[2,2]/colSums(tab)[2]
  spec=tab[1,1]/colSums(tab)[1]
  ppv=tab[2,2]/rowSums(tab)[2]
  f1=2*(sens*ppv)/(sens+ppv)

  if(is.numeric(truth)==FALSE & is.logical(truth)==FALSE){
    truth<-as.numeric(truth)-1}
  
  #CALCULATE EXACT AUC
  ord<-order(probs, decreasing=TRUE)
  probs <- probs[ord]; truth <- truth[ord]
  
  TPR=cumsum(truth)/max(1,sum(truth)) 
  FPR=cumsum(!truth)/max(1,sum(!truth))
  
  dup<-c(probs[-1]>=probs[-length(probs)], FALSE)
  TPR<-c(0,TPR[!dup],1); FPR<-c(0,FPR[!dup],1)
  
  n <- length(TPR)
  auc<- sum( ((TPR[-1]+TPR[-n])/2) * (FPR[-1]-FPR[-n]) )

  data.frame(acc,sens,spec,ppv,f1,auc)
}
class_diag(Guns$pred,Guns$y)
Guns$logit<-predict(fit2,type="link",family="binomial") 

Guns%>%ggplot()+geom_density(aes(logit,color=law,fill=law),alpha=.4)+
  theme(legend.position=c(.85,.85))+geom_vline(xintercept=0)+xlab("predictor(logit)")

ROCplot<-ggplot(Guns)+geom_roc(aes(d=y,m=prob), n.cuts=0) 
ROCplot
calc_auc(ROCplot)
```
  A logistic regression model was created to predict the presence or lack thereof a 'shall carry' law in states by prisoner incarceration rate and violent crime rates. For states with the mean prisoner incarceration rate, every unit increase in mean violent crime rates decreases the probability of having a shall carry law by 0.0049733. For states with mean violent crime rates, every unit increase in mean incarceration rates increases the probability of having a 'shall carry' law by 0.0063178. The model is doing bad with an AUC of 0.5864865. It has an accuracy of 0.7851662, sensitivity of 0.2, specificity of 0.972973, and precision of 0.7037037.
  The ROC plot of the model corresponds to an AUC value of 0.7533804 which is fair. There is about a .75 probability that a ranodmy selected state (including DC) with a shall carry law has a higher predicted probability of having the law than a randomly selected state without the shall carry law.

Logistic Regression Model - Complete
```{R}
Guns1<-read_csv("Guns.csv")%>%select(-1,-2,-state)
Guns1$y<-ifelse(Guns1$law=="yes",1,0)
Guns1<-Guns1%>%select(-law)
fit3<-glm(y~., data=Guns1,family = "binomial")
summary(fit3)
prob2<-predict(fit3,type="response")
pred2<-ifelse(prob2>.5,1,0)
table(truth=Guns$y, prediction=pred2)%>%addmargins
class_diag(pred2,Guns$y)

set.seed(6)
k=10
data<-Guns1[sample(nrow(Guns1)),]
folds<-cut(seq(1:nrow(Guns1)),breaks=k,labels=F)
diags<-NULL
for(i in 1:k){
train<-data[folds!=i,]
test<-data[folds==i,]
truth<-test$y 
fit4<-glm(y~.,data=train,family="binomial")
probs<-predict(fit4,newdata = test,type="response")
diags<-rbind(diags,class_diag(probs,truth))
}
summarize_all(diags,mean) 

y<-as.matrix(Guns1$y)
x<-model.matrix(y~.,data=Guns1)
cv <- cv.glmnet(x,y, family="binomial") 
lasso<-glmnet(x,y,family="binomial",lambda=cv$lambda.1se)

coef(lasso) #Output on RMD shows no coef for afam or murder but HTML output shows coef for all variables

set.seed(6)
Guns2<-Guns1%>%select(-afam,-murder)
k=10
data_las<-Guns2[sample(nrow(Guns2)),]
folds_las<-cut(seq(1:nrow(Guns2)),breaks=k,labels=F)
diags_las<-NULL
for(i in 1:k){
train_las<-data_las[folds_las!=i,]
test_las<-data_las[folds_las==i,]
truth_las<-test_las$y 
fit5<-glm(y~.,data=train_las,family="binomial")
probs_las<-predict(fit5,newdata = test_las,type="response")
diags_las<-rbind(diags_las,class_diag(probs_las,truth_las))
}
summarize_all(diags_las,mean) 
```
  A logistic regression model was created to predict the presence, or lack thereof, a 'shall carry' law in the US states + DC with all the variables in the dataset Guns (excluding year, state, and X1 which were classifying variables). The accuracy of the model was 0.8030691, sensitivity was 0.4210526, specificity was 0.9256757, precision was 0.6451613 to give an overall AUC of 0.6733642 which is poor. There is about a .67 probaility that a randomly selected state + DC with a shall carry law has a higher predicted probaility of having the law than a randomly selected state without the 'shall carry' law.
  A 10-fold cross validation on the model resulted in an overall better model with an AUC of 0.8291958 (pretty good!). The accuracay of the model was 0.7987976, sensitivity was 0.4141377, specificity was 0.9213843, and precision was 0.6279699. Overall, there is about a .83 probability that a randomly selected state + DC with a 'shall carry' law has a higher predicted probaility of having the law than a randomly selected state without the 'shall carry' law. The cross validation model has a slighly lower accuracy, similar specificity, similar specificity, slighly lower precision, and an overall AUC that was ~20% higher than the in-sample metrics.
  A lasso analysis on the same model retained the following variables: violent, robbery, prisoners, cauc, male, population, income, and density. Therefore, only afam and murder were eliminated. Without the afam variable being included, the 10-fold CV yeilded a AUC of 0.8179803 (good!), accuracy of 0.8022309, sensitivity of 0.4343141, specificity of 0.9194945 and precision of 0.6374604. Compared to the first 10-fold CV that included the afam and murder variable, the auc is slighly lower, but is almost the same (0.818 vs 0.829).	

